Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2022-03-26 paper/high-assurance-contracts.tex
Add minor revisions mostly to section 5

The semantics of Section \ref{agree-semantics} supports a formal
connection between the original contract---and verification results of
the AGREE model checker being run on it---and code generated from
specifications of leaf level cyber-components. It also provides the
root meaning at the base of a chain of translation steps moving from
an AGREE contract to a CakeML executables.

The first step in the chain maps the contract to a code-focused
representation. Assume given a code contract
$(I,O,A,\mathit{Eqs} \cup G \cup P,\emptyset)$, environment $E$, and
time $t$. The \emph{evaluation} of $\mathit{Eqs}$ followed by the
evaluation of $G$ results in a new environment $E'$ where the stream
values for state variables and outputs have been computed for time
$t$. The step from $E$ to $E'$ is one full cycle in the repeated
evaluation of the component.

@@

How does this notion of correctness integrate with the verification
conditions generated by AGREE? In \emph{contract correctness}, the
model checker is proving the following property
\[
\konst{Hist}(A) \imp P
\]
and in \emph{code contract correctness} we essentially prove a Hoare
triple of the form
\[
\set{\konst{Hist}(A)}\; (\mathit{Eqs} \cdot G) \; \set{P}
\]
This accords with intuition, namely that AGREE is a kind of
`program-free program logic'; by pulling out a program $(\mathit{Eqs}
\cdot G)$ from a code contract we find ourselves in the tepid
bathwater of imperative program verification. For example, at this
point one could apply a verification condition generator to help break
down the proof obligation. In order to use the code contract result in
system contract verifcation conditions, one has to show that the code
contract result implies the system verification condtion.

\subsubsection*{Wellformedness}
Stream-based evaluation allows values arbitrarily `deep in the past'
to be accessed by use of $\konst{pre}(-)$. However, the programming
languages being targeted by SPLAT do not directly support streams.  We
are currently developing a means of translating away such deep
temporal accesses by the introduction of intermediate variables. For
now, we focus on the notion of \emph{wellformed} equations that needs
to hold. The guiding intuition is that a wellformed {\Eqs} allows
access at most one step in the past, and accesses of a past value for
a variable are only possible until the variable is updated. In
informal terms, this means that ${\Eqs}\cdot G$ in a code contract
must be such that:
\begin{itemize}
\item \konst{fby} only occurs at the top level of a variable definition, \eg, $v = e_1 \to e2$;
\item $\konst{pre}(e)$ is allowed only when $e$ is a variable; and
\item $\konst{pre}(v)$ is allowed only before the defining equation $v = e$ (and may occur in $e$).
\end{itemize}
For example, the Fibonacci sequence $1,1,2,3,5,8,13,\ldots$ can be
defined, albeit in a non-wellformed manner, by
{\small
\begin{verbatim}
  Fib = 1 -> pre(1 -> Fib + pre Fib)
\end{verbatim}
}
\noindent or by (again non-wellformed)
{\small
\begin{verbatim}
  N = 0 -> 1 + pre(N)
  Fib = if N <= 1 then 1 else pre(Fib) + pre(pre(Fib)).
\end{verbatim}
}
\noindent or by (wellformed):
{\small
\begin{verbatim}
  N = 0 -> 1 + pre(N)
  A = 42 -> pre(Fib)
  Fib = if N <= 1 then 1 else pre(Fib) + pre(A)
\end{verbatim}
}
A wellformed ${\Eqs} \cdot G$ is straightforward to implement in
conventional programming languages, as we will see in
Section \ref{sec:synthesis}.

@@

\newsavebox{\flt}
\begin{lrbox}{\flt}
  \begin{lstlisting}[style=agree,numbers=left] -- start user
  definitions eq policy : bool =
  WELL_FORMED_AUTOMATION_RESPONSE(Input); -- stop user definitions

@@

Code contracts for the filter and monitor are shown in
\figref{fig:filter} and \figref{fig:monitor}.\footnote{KLS: do we discuss wellformedness anywhere?} In general, the
\konst{eq}-statements of a code contract keep track of
state and provide data values from which to compute values to place on
the outputs.

\subsubsection{Monitor specification}
BriefCASE generates everything but the
\konst{eq}-statements in \linesref{line:mon-assume}{line:mon-policy} of
\figref{fig:monitor}.  As with the filter, the guarantees
completely define the output, as required for synthesis, in terms of
the policy and the alert.  The
\texttt{alert} variable defined on \lineref{line:mon-alert} depends
not just on the policy but also on configuration details supplied when
the component is specified.  At configuration
time, \texttt{is\_latched} has been set to \konst{true}, thus making
the alert persistent, meaning that once the alert is raised, it is
always raised; otherwise, it is the complement of the policy value in
the current step.

@@

The latency bound for the monitor is defined by the constant
\texttt{MAX\_LATENCY} on \lineref{line:mon-const}.  Here the constant
is set to one to be consistent with the system specification from the
previous section.  At time 0 the policy is trivial: a response
requires an accompanying request.  After time 0, if a request is
pending in the current time step, then the latency must still be under
the bound, and if there is a response, then it closes a request now or
a pending request from earlier in time.

>>>>

The semantics of Section \ref{agree-semantics} supports a formal
connection between the original contract---and verification results of
the AGREE model checker being run on it---and code generated from
specifications of leaf level cyber-components. It also provides the
root meaning at the base of a chain of translation steps moving from
an AGREE contract to a CakeML executable.

The first step in the chain maps the contract to a code-focused
representation. Assume a code contract
$(I,O,A,\mathit{Eqs} \cup G \cup P,\emptyset)$, environment $E$, and
time $t$. The \emph{evaluation} of $\mathit{Eqs}$ followed by the
evaluation of $G$ results in a new environment $E'$ where the stream
values for state variables and outputs have been computed for time
$t$. The step from $E$ to $E'$ is one full cycle in the repeated
evaluation of the component.

@@

How does this notion of correctness integrate with the verification
conditions generated by AGREE? In \emph{contract correctness}, the
model checker is proving the following property
\[
\konst{Hist}(A) \imp P
\]
and in \emph{code contract correctness} we essentially prove a Hoare
triple of the form
\[
\set{\konst{Hist}(A)}\; (\mathit{Eqs} \cdot G) \; \set{P}
\]
This accords with intuition, namely that AGREE is a kind of
`program-free program logic'; by pulling out a program $(\mathit{Eqs}
\cdot G)$ from a code contract we find ourselves in the domain of imperative program verification. For example, at this
point one could apply a verification condition generator to help break
down the proof obligation. In order to use the code contract result in
system contract verification conditions, one has to show that the code
contract result implies the system verification condition.

\subsubsection*{Wellformedness}
Stream-based evaluation allows values arbitrarily `deep in the past'
to be accessed by use of $\konst{pre}(-)$. However, the programming
languages being targeted by SPLAT do not directly support streams.  We
are currently developing a means of translating away such deep
temporal accesses by the introduction of intermediate variables. For
now, we focus on the notion of \emph{wellformed} equations that needs
to hold. The guiding intuition is that a wellformed {\Eqs} allows
access at most one step in the past, and accesses of a past value for
a variable are only possible until the variable is updated. In
informal terms, this means that ${\Eqs}\cdot G$ in a code contract
must be such that:
\begin{itemize}
\item Followed-by only occurs at the top level of a variable definition, \eg, $v = e_1 \to e2$;
\item $\konst{pre}(e)$ is allowed only when $e$ is a variable; and
\item $\konst{pre}(v)$ is allowed only before the defining equation $v = e$ (and may occur in $e$).
\end{itemize}
For example, the Fibonacci sequence $1,1,2,3,5,8,13,\ldots$ can be
defined, albeit in a non-wellformed manner, by
{\small
\begin{verbatim}
  Fib = 1 -> pre(1 -> Fib + pre Fib)
\end{verbatim}
}
\noindent or by (again non-wellformed)
{\small
\begin{verbatim}
  N = 0 -> 1 + pre(N)
  Fib = if N <= 1 then 1 else pre(Fib) + pre(pre(Fib)).
\end{verbatim}
}
\noindent or by (wellformed):
{\small
\begin{verbatim}
  N = 0 -> 1 + pre(N)
  A = 42 -> pre(Fib)
  Fib = if N <= 1 then 1 else pre(Fib) + pre(A)
\end{verbatim}
}
A wellformed ${\Eqs} \cdot G$ is straightforward to implement in
conventional programming languages, as we will see in
Section \ref{sec:synthesis}.

@@

\newsavebox{\flt}
\begin{lrbox}{\flt}
  \begin{lstlisting}[style=agree,numbers=left] -- start user
    -- start user definitions 
    eq policy : bool =
      WELL_FORMED_AUTOMATION_RESPONSE(Input); 
    -- stop user definitions

@@

Code contracts for the filter and monitor are shown in
\figref{fig:filter} and \figref{fig:monitor}.
The filter is direct: it defines the policy according to the well-formed requirement for data.
We now discuss in some detail the monitor.

\subsubsection{Monitor specification}
BriefCASE generates everything but the
\konst{eq}-statements in \linesref{line:mon-assume}{line:mon-policy} of
\figref{fig:monitor}. 
The
\texttt{alert} variable defined on \lineref{line:mon-alert} depends
not just on the policy but also on configuration details supplied when
the component is specified.  At configuration
time, \texttt{is\_latched} has been set to \konst{true}, thus making
the alert persistent, meaning that once the alert is raised, it is
always raised; otherwise, it is the complement of the policy value in
the current step.

@@

The latency bound for the monitor is defined by the constant
\texttt{MAX\_LATENCY} on \lineref{line:mon-const}.  Here the constant
is set to one to be consistent with the system specification from the
previous section.  At time zero the policy is trivial: a response
requires an accompanying request.  After time zero, if a request is
pending in the current time step, then the latency must still be under
the bound, and if there is a response, then it closes a request now or
a pending request from earlier in time.

### 2022-03-26 paper/sw-contract.tex
Add minor revisions mostly to section 5

The assumption on \lineref{line:sw-assume-3} constrains when a request can arrive by reasoning about the \emph{state} of the contract.
The state of the contract is captured by stream semantics of AGREE.
When writing assumptions and guarantees, it is important to differentiate pre-state, before a contract updates its state, and post-state, after a contract updates its state, in response to the current input.
The \texttt{pre} naturally makes that distinction.
In general, assumptions regarding state should be evaluated in the pre-state of the component with the current inputs, and guarantees regarding state should be evaluated on the post-state of the component given the current inputs.
Guarantees should use \texttt{pre} anytime they need to reason about the current state relative to the previous state.

The assumption on \lineref{line:sw-assume-3} uses a followed-by expression so it is \texttt{true} at time 0 and then it is the truth value of the implication in all future instances.
The followed-by guards the \texttt{pre} as mentioned previously;
thus after time 0, the assumption depends on the presence of a request and the pre-state of the component.
The assumption on \lineref{line:sw-assume-3} is that if there is an incoming request, it is either the very first one, \texttt{\textbf{Historically}(\textbf{not} req)}, or it is after the component has output a start event in response to a previous request, \texttt{\textbf{Since}(\textbf{not} req, strt)}--\emph{not request since start}.

The guarantees on \linesref{line:sw-guarantee-1}{line:sw-guarantee-3} coincide events and assert well-formed output.
The guarantees on \linesref{line:sw-guarantee-4}{line:sw-guarantee-6} define temporal properties of the component.
\lineref{line:sw-guarantee-4} insists that a start happens with a request or one step after a request.
The guarantee uses the assumption on \lineref{line:sw-assume-3} and does not check for two requests in a row without a start as the assumption precludes that input behavior.
It differentiates with the followed-by what is required at time 0, the start must coincide with the request, with what is required after time 0, the start must coincide with the request or is a response to a request one step earlier.
The guarantee also does not force the start to always happen, it only says that if it does happen, it is under the defined conditions.

>>>>

The assumption on \lineref{line:sw-assume-3} constrains when a request can arrive by reasoning about the \emph{state} of the contract defined be the associated streams.
When writing assumptions and guarantees, it is important to differentiate pre-state, before a contract updates its state, and post-state, after a contract updates its state, in response to the current input.
The \texttt{pre} naturally makes that distinction.
In general, assumptions regarding state should be evaluated in the pre-state of the component with the current inputs, and guarantees regarding state should be evaluated on the post-state of the component given the current inputs.
Additionally, guarantees should use \texttt{pre} anytime they need to reason about the current state relative to the previous state.

The assumption on \lineref{line:sw-assume-3} uses a followed-by expression so it is \texttt{true} at time zero and then it is the truth value of the implication in all future instances.
The followed-by guards the \texttt{pre} as mentioned previously;
thus after time zero, the assumption depends on the presence of a request and the pre-state of the component.
The assumption on \lineref{line:sw-assume-3} is that if there is an incoming request, it is either the very first one, \texttt{\textbf{Historically}(\textbf{not} req)}, or it is after the component has output a start event in response to a previous request, \texttt{\textbf{Since}(\textbf{not} req, strt)}---\emph{not request since start}.

The guarantees on \linesref{line:sw-guarantee-1}{line:sw-guarantee-3} coincide events and assert well-formed output.
The guarantees on \linesref{line:sw-guarantee-4}{line:sw-guarantee-6} define temporal properties of the component.
\lineref{line:sw-guarantee-4} insists that a start happens with a request or one step after a request.
The guarantee uses the assumption on \lineref{line:sw-assume-3} and does not check for two requests in a row without a start as the assumption precludes that input behavior.
It differentiates with the followed-by what is required at time zero, the start must coincide with the request, with what is required after time zero, the start must coincide with the request or is a response to a request one step earlier.
The guarantee also does not force the start to always happen, it only says that if it does happen, it is under the defined conditions.
